Appendix: Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-04-02 Appendix_plain_v3.tex
update

\subsection*{Bioinformatics}
Starting with Illumina-demultiplexed sequencing files, we first removed primer and adapter sequences using Cutadapt v4.9, applying the default mismatch tolerance parameter (-e 0.1) \cite{martin2011}. Next, we performed quality filtering using the filterAndTrim function from the dada2 R package, retaining its default parameters while specifying the optional truncLen argument to match the minimum MiFish amplicon length. After filtering, we denoised the sequences using dada2 with default settings \cite{callahan2016}, which resulted in a list of Amplicon Sequence Variants (ASVs) across all samples. Each resulting ASV was then assigned a unique 28-character hash code using the sha-1 algorithm with the R function map\_chr. This allowed us to create a common database of ASVs by hash, enabling us to track which sequences have already been annotated for consistency and to avoid redundant computational efforts.

@@

The above statistical model provides predictions of species-specific DNA concentrations ($E$) at each depth sampled (d = 0, 50, 150, 300 or 500 m). To generate smooth distribution models, we use posterior means of $E_s(x,y,d)$ from the joint statistical model (SI Appendix Figure S.\ref{fig:DAG_v3}), exponentiated it to move to concentration from log concentration, $G_{s}(x,y,d) = exp(E_{s}(x,y,d))$, and used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA. Because of the predominance of very low estimated DNA concentrations at a large number of site, we set $G_{s}(x,y,d) < 0.5$ equal to zero (SI Appendix Figure S.\ref{fig:presence_threshold}) and then used a Tweedie likelihood in sdmTMB,

@@

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/DAG_v3.pdf} 
\caption{Schematic for the statistical models to produce species distributions. Three sets of observations (orange) were used simultaneously to provide estimates of eDNA concentration for  twelve fish species (blue) from which we derive smoothed predictions of DNA concentration (green). Mock communities for metabarcoding (yellow) were estimated externally and provided estimated $\alpha_s$ that were used in the joint model. Notation for parameters, subscripts, and observations are indicated in gray.}
\label{fig:DAG_v3}
\end{figure*}

>>>>

\subsection*{Bioinformatics}
Starting with Illumina-demultiplexed sequencing files, we first removed primer and adapter sequences using Cutadapt v4.9, applying the default mismatch tolerance parameter (-e 0.1) \cite{martin2011}. Next, we performed quality filtering using the filterAndTrim function from the DADA2 R package, retaining its default parameters while specifying the optional truncLen argument to match the minimum MiFish amplicon length. After filtering, we denoised the sequences using DADA2 with default settings \cite{callahan2016}, which resulted in a list of Amplicon Sequence Variants (ASVs) across all samples. Each resulting ASV was then assigned a unique 28-character hash code using the sha-1 algorithm with the R function map\_chr. This allowed us to create a common database of ASVs by hash, enabling us to track which sequences have already been annotated for consistency and to avoid redundant computational efforts.

@@

The above statistical model provides predictions of species-specific DNA concentrations ($E$) at each depth sampled (d = 0, 50, 150, 300 or 500 m). To generate smooth distribution models, we use posterior means of $E_s(x,y,d)$ from the joint statistical model, exponentiated it to move to concentration from log concentration, $G_{s}(x,y,d) = exp(E_{s}(x,y,d))$, and used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA. Because of the predominance of very low estimated DNA concentrations at a large number of site, we set $G_{s}(x,y,d) < 0.5$ equal to zero (SI Appendix Figure S.\ref{fig:presence_threshold}) and then used a Tweedie likelihood in sdmTMB,

@@

%\begin{figure*}[tbhp] 
%\centering
%\includegraphics[width=16cm]{plots/DAG_v3.pdf} 
%\caption{Schematic for the statistical models to produce species distributions. Three sets of observations (orange) were used simultaneously to provide estimates of eDNA concentration for  twelve fish species (blue) from which we derive smoothed predictions of DNA concentration (green). Mock communities for metabarcoding (yellow) were estimated externally and provided estimated $\alpha_s$ that were used in the joint model. Notation for parameters, subscripts, and observations are indicated in gray.}
%\label{fig:DAG_v3}
%\end{figure*}
